---
title: Analysis of CFTR variants
author: Elena Núñez Tomás
output: html_document
email_address: elena.nt.bioinf@gmail.com
github_repo: https://github.com/elenantbioinf/poc_cf
params:
    final_report_json: NULL
    qc_summary_tsv: NULL
    raw_multiqc_report: NULL
    clean_multiqc_report: NULL
    target_region_evaluability_plot: NULL
    coverage_gaps_tsv: NULL
    gap_min_coverage: NULL
    target_bed: NULL
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Montserrat:ital,wght@0,100..900;1,100..900&display=swap" rel="stylesheet">


```{css echo=FALSE}
.author, .title {
    display: none;
}


h1 {
    text-align: center;
    background-color: #d2d9e7;
    padding: 25px;
    border-bottom: 3px solid #1a1a1a;
    margin-bottom: 40px;
    border-radius: 6px;
}

h2 {
    padding: 25px;
    background-color: #ebf0f8;
    border-bottom: 3px solid #1a1a1a;
    margin-bottom: 40px;
    border-radius: 6px;
}

body {
    background-color: white;
    color: black;
    font-family: "Montserrat", sans-serif;
    font-size: 18px;
}
pre {
    background-color: #f4f4f4;
    padding: 15px;
    border-radius: 6px;
}

.main-container{
    max-width: 1000px;
    margin: 60px auto;
    padding: 50px;
    background-color: white;
    border: 1px solid #dcdcdc;
    border-radius: 10px;
    box-shadow: 0 4px 12px rgba(0,0,0,0.05);
}

.report-subtitle{
  text-align: center;
  margin-top: -25px;
  margin-bottom: 30px;
  color: #333333;
  font-size: 16px;
  font-weight: 400;
}

.report-description{
    text-align: center;
    max-width: 780px;
    margin: 0 auto 45px auto;
    color: #333333;
    font-size: 20px;
    line-height: 1.5;
}

.report-metadata{
    max-width: 780px;
    margin: 0 auto 45px auto;
    padding: 18px 24px;
    background-color: #f5f5f5;
    border: 1px solid #ebedf0;
    border-radius: 8px;
    box-shadow: 0 3px 10px rgba(0,0,0,0.06);
    color: #4b4b4b;
    font-size: 16px;
    line-height: 1.5;
}

.report-metadata p {
    margin: 5px 0;
}

.report-metadata a {
    color: #4b4b4b;
    text-decoration: none;
}

.report-metadata a:hover {
    text-decoration: underline;
}

.analysis-summary{
    max-width: 780px;
    margin: 0 auto 45px auto;
    padding: 18px 24px;
    background-color: #f8f8f8;
    border-left: 10px solid #1f3a5f;
    border-radius: 8px;
    box-shadow: 0 2px 6px rgba(0,0,0,0.04);
    color: #333333;
    font-size: 16px;
    line-height: 1.5;
}

.analysis-summary p {
    margin: 5px 0;
}

.analysis-summary .interpretation {
    margin-top: 12px;
}

.technical-limitations{
    max-width: 780px;
    margin: 0 auto 45px auto;
    padding: 18px 24px;
    background-color: #f8f8f8;
    border-left: 10px solid #832121;
    border-radius: 8px;
    box-shadow: 0 2px 6px rgba(0, 0, 0, 0.04);
    color: #333333;
    font-size: 16px;
    line-height: 1.5;
}

.technical-limitations p {
    margin: 5px 0;
}

.sample_information{
    max-width: 780px;
    margin: 0 auto 45px auto;
    padding: 18px 24px;
    background-color: #f8f8f8;
    border-left: 10px solid #1f3a5f;
    border-radius: 8px;
    box-shadow: 0 2px 6px rgba(0,0,0,0.04);
    color: #333333;
    font-size: 20px;
    line-height: 1.5;
}

.coverage-summary {
    max-width: 780px;
    margin: 20px auto 45px auto;
    padding: 18px 22px;
    border-radius: 6px;
    background-color: #f8f8f8;
    box-shadow: 0 2px 6px rgba(0, 0, 0, 0.04);
    font-size: 16px;
    line-height: 1.5;
    box-sizing: border-box;
}

.coverage-summary p {
    margin: 5px 0;
}

.coverage-ok {
    border-left: 10px solid #4d7f4d;
}

.coverage-warning {
    border-left: 10px solid #832121;
}

.coverage-interpretation{
    max-width: 780px;
    margin: 20px auto 45px auto;
    padding: 18px 24px;
    background-color: #f8f8f8;
    border-left: 10px solid #1f3a5f;
    border-radius: 8px;
    box-shadow: 0 2px 6px rgba(0, 0, 0, 0.04);
    color: #333333;
    font-size: 14px;
    line-height: 1.5;
}

.coverage-interpretation p {
    margin: 5px 0;
}

.coverage-green {color: #4d7f4d;font-weight: 700}

.coverage-red {color: #832121;font-weight: 700}

.qc-summary-table {
    max-width: 900px;
    margin: 20px auto 45px auto;
    overflow-x: auto;
}

.qc-summary-table table {
    width: 100%;
    border-collapse: collapse;
    font-size: 15px;
}

.qc-summary-table th {
    background-color: #d2d9e7;
    border: 1px solid #c4c8cf;
    padding: 10px;
    text-align: center;
}

.qc-summary-table td {
    border: 1px solid #dcdcdc;
    padding: 10px;
    text-align: center;
}

.qc-summary-table tr:nth-child(even) {
    background-color: #f8f8f8;
}

.qc-interpretation {
    max-width: 780px;
    margin: 20px auto 45px auto;
    padding: 18px 24px;
    background-color: #f8f8f8;
    border-left: 10px solid #1f3a5f;
    border-radius: 8px;
    box-shadow: 0 2px 6px rgba(0,0,0,0.04);
    color: #333333;
    font-size: 15px;
    line-height: 1.5;
}
```

```{r echo=FALSE}
library(jsonlite)

data <- fromJSON(params$final_report_json)

sample_id <- data$sample_id
disease   <- data$disease
gene      <- data$gene
reference    <- data$reference_genome

qc_summary_tsv <- params$qc_summary_tsv
raw_multiqc_report <- params$raw_multiqc_report
clean_multiqc_report <- params$clean_multiqc_report
target_bed <- params$target_bed
target_region_evaluability_plot <- params$target_region_evaluability_plot
coverage_gaps_tsv <- params$coverage_gaps_tsv
gap_min_coverage <- params$gap_min_coverage

total_variants <- data$summary$total_unique_variants
variants <- data$variants
no_variants <- (total_variants == 0)

report_author <- "Elena Núñez Tomás"
report_email <- "elena.nt.bioinf@gmail.com"
report_generation_date <- format(Sys.time(), "%Y-%m-%d %H:%M:%S %Z")
github_repo <- "https://github.com/elenantbioinf/poc_cf"
```

# `r gene` Variant and Coverage Evaluability Report

<div class="report-subtitle">
Automated proof-of-concept report generated by the Snakemake pipeline
</div>

<div class="report-description">
This report integrates variant analysis, functional annotation and target-region coverage evaluability for `r gene` in the context of `r disease`. It is designed to distinguish detected variants from regions that were technically assessed according to the configured coverage threshold.
</div>

<div class="report-metadata">
  <p><strong>Pipeline created by:</strong> `r report_author`</p>
  <p><strong>GitHub repository:</strong> <a href="`r github_repo`" target="_blank">`r github_repo`</a></p>
  <p><strong>Contact:</strong> <a href="mailto:`r report_email`">`r report_email`</a></p>
  <p><strong>Report generated on:</strong> `r report_generation_date`</p>
</div>

## Analysis summary

```{r analysis-summary, echo=FALSE, results="asis", warning=FALSE}
format_percent <- function(value) {
  sprintf("%.2f%%", value)
}

analysis_gap_count <- NA_integer_
analysis_total_target_bp <- NA_integer_
analysis_total_low_coverage_bp <- NA_integer_
analysis_low_coverage_bp_percent <- NA_real_
analysis_affected_target_intervals <- NA_integer_
analysis_total_target_intervals <- NA_integer_
analysis_affected_target_intervals_percent <- NA_real_

if (
  !is.null(coverage_gaps_tsv) && file.exists(coverage_gaps_tsv) &&
  !is.null(target_bed) && file.exists(target_bed)
) {

  gaps_df <- read.delim(
    coverage_gaps_tsv,
    sep = "\t",
    stringsAsFactors = FALSE
  )

  bed_df <- read.delim(
    target_bed,
    sep = "\t",
    header = FALSE,
    stringsAsFactors = FALSE,
    comment.char = "#"
  )

  colnames(bed_df)[1:3] <- c("chrom", "start", "end")

  bed_df$start <- as.integer(bed_df$start)
  bed_df$end <- as.integer(bed_df$end)

  analysis_gap_count <- nrow(gaps_df)
  analysis_total_target_bp <- sum(bed_df$end - bed_df$start)
  analysis_total_target_intervals <- nrow(bed_df)

  if (nrow(gaps_df) > 0) {
    gaps_df$start <- as.integer(gaps_df$start)
    gaps_df$end <- as.integer(gaps_df$end)
    gaps_df$length <- as.integer(gaps_df$length)

    analysis_total_low_coverage_bp <- sum(gaps_df$length, na.rm = TRUE)
    analysis_low_coverage_bp_percent <- (
      analysis_total_low_coverage_bp / analysis_total_target_bp
    ) * 100

    affected <- vapply(
      seq_len(nrow(bed_df)),
      function(index) {
        any(
          gaps_df$chrom == bed_df$chrom[index] &
            gaps_df$end > bed_df$start[index] &
            gaps_df$start < bed_df$end[index]
        )
      },
      logical(1)
    )

    analysis_affected_target_intervals <- sum(affected)
    analysis_affected_target_intervals_percent <- (
      analysis_affected_target_intervals / analysis_total_target_intervals
    ) * 100
  } else {
    analysis_total_low_coverage_bp <- 0
    analysis_low_coverage_bp_percent <- 0
    analysis_affected_target_intervals <- 0
    analysis_affected_target_intervals_percent <- 0
  }
}

if (is.na(analysis_gap_count)) {

  cat(
    paste0(
      "\n<div class='analysis-summary'>",
      "<p><strong>Detected variants:</strong> ", total_variants, "</p>",
      "<p><strong>Coverage summary:</strong> not available.</p>",
      "<p class='interpretation'><strong>Interpretation:</strong> variant findings should be interpreted together with the available quality-control and coverage information.</p>",
      "</div>\n"
    )
  )

} else {

  if (total_variants == 0 && analysis_gap_count > 0) {
    interpretation_text <- paste0(
      "No variants were detected in the analysed data. However, ",
      "low-coverage target regions were identified, so this result should ",
      "be interpreted together with the coverage evaluability assessment."
    )
  } else if (total_variants == 0 && analysis_gap_count == 0) {
    interpretation_text <- paste0(
      "No variants were detected and no low-coverage gaps were identified ",
      "below the configured coverage threshold."
    )
  } else {
    interpretation_text <- paste0(
      "Detected variants should be interpreted together with functional ",
      "annotation and target-region coverage evaluability."
    )
  }

  cat(
    paste0(
      "\n<div class='analysis-summary'>",
      "<p><strong>Detected variants:</strong> ", total_variants, "</p>",
      "<p><strong>Coverage threshold:</strong> ", gap_min_coverage, "X</p>",
      "<p><strong>Low-coverage gaps detected:</strong> ", analysis_gap_count, "</p>",
      "<p><strong>Non-evaluable target bases at ", gap_min_coverage, "X:</strong> ",
      format_percent(analysis_low_coverage_bp_percent), "</p>",
      "<p><strong>Target intervals affected by low coverage:</strong> ",
      analysis_affected_target_intervals, " / ", analysis_total_target_intervals,
      " (", format_percent(analysis_affected_target_intervals_percent), ")</p>",
      "<p class='interpretation'><strong>Interpretation:</strong> ",
      interpretation_text,
      "</p>",
      "</div>\n"
    )
  )
}
```

## Sample information

<div class="sample_information">

<strong>Sample ID:</strong> `r sample_id`

<strong>Disease:</strong> `r disease`

<strong>Gene:</strong> `r gene`

<strong>Reference genome:</strong> `r reference`

</div>

## Sequencing quality summary

This table compares the FastQC metrics reported by MultiQC before and after read preprocessing.

```{r qc-summary-table, echo=FALSE, results="asis", warning=FALSE}
if (!is.null(qc_summary_tsv) && file.exists(qc_summary_tsv)) {

  qc_df <- read.delim(
    qc_summary_tsv,
    sep = "\t",
    stringsAsFactors = FALSE,
    check.names = FALSE
  )

  display_qc <- qc_df[
    ,
    c(
      "read",
      "stage",
      "total_sequences",
      "sequence_length",
      "avg_sequence_length",
      "percent_gc",
      "percent_duplicates"
    )
  ]
   
  colnames(display_qc) <- c(
    "Read",
    "Stage",
    "Total sequences",
    "Sequence length",
    "Average sequence length",
    "GC (%)",
    "Duplicated sequences (%)"
  )

  display_qc[["Total sequences"]] <- format(
    display_qc[["Total sequences"]],
    big.mark = ",",
    scientific = FALSE
  )

  display_qc[["Average sequence length"]] <- sprintf(
    "%.2f",
    as.numeric(display_qc[["Average sequence length"]])
  )

  display_qc[["GC (%)"]] <- sprintf(
    "%.2f",
    as.numeric(display_qc[["GC (%)"]])
  )

  display_qc[["Duplicated sequences (%)"]] <- sprintf(
    "%.2f",
    as.numeric(display_qc[["Duplicated sequences (%)"]])
  )

  cat("<div class='qc-summary-table'>\n")

  cat(
    knitr::kable(
      display_qc,
      format = "html",
      escape = FALSE,
      row.names = FALSE
    )
  )
  cat("</div>\n")

} else {
  cat("QC summary table was not available.\n")
}
```

```{r qc-summary-interpretation, echo=FALSE, results="asis", warning=FALSE}
if (
  exists("qc_df") &&
  nrow(qc_df) > 0 &&
  all(
    c(
      "read",
      "stage",
      "total_sequences",
      "avg_sequence_length",
      "percent_gc",
      "percent_duplicates"
    ) %in% colnames(qc_df)
  )
) {

  raw_qc <- qc_df[qc_df$stage == "Raw", ]
  filtered_qc <- qc_df[qc_df$stage == "Post-filtering", ]

  common_reads <- intersect(
    raw_qc$read,
    filtered_qc$read
  )

  if (length(common_reads) > 0) {

    retained_percentages <- vapply(
      common_reads,
      function(read_name) {
        raw_total <- as.numeric(
          raw_qc$total_sequences[raw_qc$read == read_name][1]
        )

        filtered_total <- as.numeric(
          filtered_qc$total_sequences[filtered_qc$read == read_name][1]
        )

        (filtered_total / raw_total) * 100
      },
      numeric(1)
    )

    retained_percent <- mean(
      retained_percentages,
      na.rm = TRUE
    )

    raw_mean_length <- mean(
      as.numeric(raw_qc$avg_sequence_length),
      na.rm = TRUE
    )

    filtered_mean_length <- mean(
      as.numeric(filtered_qc$avg_sequence_length),
      na.rm = TRUE
    )

    mean_length_change <- filtered_mean_length - raw_mean_length

    raw_gc <- mean(
      as.numeric(raw_qc$percent_gc),
      na.rm = TRUE
    )

    filtered_gc <- mean(
      as.numeric(filtered_qc$percent_gc),
      na.rm = TRUE
    )

    raw_duplicates <- mean(
      as.numeric(raw_qc$percent_duplicates),
      na.rm = TRUE
    )

    filtered_duplicates <- mean(
      as.numeric(filtered_qc$percent_duplicates),
      na.rm = TRUE
    )

    cat(
      paste0(
        "\n<div class='qc-interpretation'>",
        "<p><strong>Interpretation:</strong> ",
        sprintf("%.2f%%", retained_percent),
        " of reads were retained after preprocessing. ",
        "The average sequence length changed from ",
        sprintf("%.2f", raw_mean_length),
        " bp to ",
        sprintf("%.2f", filtered_mean_length),
        " bp (",
        sprintf("%+.2f", mean_length_change),
        " bp). ",
        "The mean GC content remained stable at approximately ",
        sprintf("%.2f%%", filtered_gc),
        ". ",
        "The estimated duplicated-sequence percentage changed from ",
        sprintf("%.2f%%", raw_duplicates),
        " to ",
        sprintf("%.2f%%", filtered_duplicates),
        ".</p>",
        "</div>\n"
      )
    )
  }

} else {
  cat(
    "\n<div class='qc-interpretation'>",
    "<p><strong>Interpretation:</strong> ",
    "The sequencing quality metrics could not be interpreted automatically.",
    "</p>",
    "</div>\n"
  )
}
```

<div class="qc-report-links">

<p><strong>Detailed quality-control reports:</strong></p>

<p>
<a href="`r raw_multiqc_report`" target="_blank">
Open raw-read MultiQC report
</a>
</p>

<p>
<a href="`r clean_multiqc_report`" target="_blank">
Open post-filtering MultiQC report
</a>
</p>

</div>

## Target region coverage evaluability

This figure summarizes which target BED intervals were technically evaluable according to the minimum coverage threshold used by the pipeline.

```{r target-region-evaluability, echo=FALSE, results="asis", warning=FALSE}
if (!is.null(target_region_evaluability_plot) && file.exists(target_region_evaluability_plot)) {
  image_src <- knitr::image_uri(target_region_evaluability_plot)

  cat(
    sprintf(
      '<img src="%s" alt="Target region coverage evaluability map" style="width:100%%; display:block; margin:20px auto;">',
      image_src
    )
  )
} else {
  cat("Target region evaluability plot was not available.\n")
}
```

<div class="coverage-interpretation">

<p><strong><center>How to read this figure</center></strong></p>

<p>Each block represents one `r gene` target interval included in the BED file.</p>

<p><span class="coverage-green">Green regions</span> indicate target bases with coverage equal to or above `r gap_min_coverage`X.</p>

<p><span class="coverage-red">Red regions</span> indicate low-coverage bases below `r gap_min_coverage`X and should be considered technically non-evaluable at this threshold.</p>

<p>The plot is intended as an evaluability overview. It does not represent the physical genomic distance between target intervals.</p>

</div>

### Coverage gap summary

```{r coverage-gap-summary, echo=FALSE, results="asis", warning=FALSE}
format_bp <- function(value) {
  format(value, big.mark = ",", scientific = FALSE)
}

format_percent <- function(value) {
  sprintf("%.2f%%", value)
}

read_target_bed <- function(target_bed_file) {
  bed_df <- read.delim(
    target_bed_file,
    sep = "\t",
    header = FALSE,
    stringsAsFactors = FALSE,
    comment.char = "#"
  )

  if (ncol(bed_df) < 3) {
    stop("Target BED file must contain at least three columns: chrom, start, end.")
  }

  colnames(bed_df)[1:3] <- c("chrom", "start", "end")

  bed_df$start <- as.integer(bed_df$start)
  bed_df$end <- as.integer(bed_df$end)

  return(bed_df)
}

count_affected_target_intervals <- function(bed_df, gaps_df) {
  affected <- vapply(
    seq_len(nrow(bed_df)),
    function(index) {
      any(
        gaps_df$chrom == bed_df$chrom[index] &
          gaps_df$end > bed_df$start[index] &
          gaps_df$start < bed_df$end[index]
      )
    },
    logical(1)
  )

  sum(affected)
}

if (
  !is.null(coverage_gaps_tsv) && file.exists(coverage_gaps_tsv) &&
  !is.null(target_bed) && file.exists(target_bed)
) {

  gaps_df <- read.delim(
    coverage_gaps_tsv,
    sep = "\t",
    stringsAsFactors = FALSE
  )

  bed_df <- read_target_bed(target_bed)

  total_target_bp <- sum(bed_df$end - bed_df$start)
  total_target_intervals <- nrow(bed_df)

  if (nrow(gaps_df) == 0) {

    cat(
      paste0(
        "\n<div class='coverage-summary coverage-ok'>",
        "<p>No low-coverage gaps were detected below ", gap_min_coverage, "X.</p>",
        "<p>All target BED intervals are technically evaluable according to this coverage criterion.</p>",
        "<p>Total target length assessed: ", format_bp(total_target_bp), " bp.</p>",
        "</div>\n"
      )
    )

  } else {

    gaps_df$start <- as.integer(gaps_df$start)
    gaps_df$end <- as.integer(gaps_df$end)
    gaps_df$length <- as.integer(gaps_df$length)

    total_low_coverage_bp <- sum(gaps_df$length, na.rm = TRUE)
    low_coverage_bp_percent <- (total_low_coverage_bp / total_target_bp) * 100

    affected_target_intervals <- count_affected_target_intervals(
      bed_df = bed_df,
      gaps_df = gaps_df
    )

    affected_target_intervals_percent <- (
      affected_target_intervals / total_target_intervals
    ) * 100

    cat(
      paste0(
        "\n<div class='coverage-summary coverage-warning'>",
        "<p>Low-coverage gaps detected: ", nrow(gaps_df), "</p>",
        "<p>Minimum coverage threshold: ", gap_min_coverage, "X</p>",
        "<p>Total target length assessed: ", format_bp(total_target_bp), " bp</p>",
        "<p>Total low-coverage length: ", format_bp(total_low_coverage_bp), " bp</p>",
        "<p>Non-evaluable target bases: ", format_percent(low_coverage_bp_percent), "</p>",
        "<p>Target intervals affected by low coverage: ", affected_target_intervals,
        " / ", total_target_intervals,
        " (", format_percent(affected_target_intervals_percent), ")</p>",
        "<p>Regions overlapping these gaps should be considered technically non-evaluable for variant interpretation at this threshold.</p>",
        "</div>\n"
      )
    )
  }

} else {
  cat("Coverage gap summary was not available.\n")
}
```

## Variants analysis

Total unique variants: `r total_variants`

```{r echo=FALSE, results="asis", warning=FALSE}
if (no_variants) {
  cat("No variants were detected in the analysed data. However, this result should be interpreted together with the coverage evaluability assessment, since low-coverage target regions may limit the ability to detect variants in those intervals.\n")
} else {
  for (i in seq_len(nrow(variants))) {
    v <- variants[i, ]

    cat(sprintf("\n### Variant: %s\n\n", v$input_variant))
    cat(sprintf("- Best annotation: %s\n", v$best_annotation))
    cat(sprintf("- Gene: %s\n", v$gene))
    cat(sprintf("- Transcript: %s\n", v$transcript))
    cat(sprintf("- Exon: %s\n", v$exon))
    cat(sprintf("- Consequence: %s\n", v$consequence))
    cat(sprintf("- Impact: %s\n", v$impact))
    cat(sprintf("- HGVS_C: %s\n", v$hgvs_c))
    cat(sprintf("- Protein: %s\n", v$protein))
    cat(sprintf("- Transcript annotations (rows): %s\n\n", v$transcript_annotations_rows))
  }
}
```

## Technical limitations

<div class="technical-limitations">

<p>
This report is generated as a proof-of-concept and should not be considered a validated diagnostic report.
</p>

<p>
Variant interpretation is limited by the sequencing data, target-region coverage, reference genome, annotation sources and configured filtering criteria.
</p>

<p>
Regions classified as non-evaluable due to low coverage may limit the ability to detect variants in those intervals.
</p>

</div>