---
title: Analysis of CFTR variants
author: Elena Núñez Tomás
output: html_document
email_address: elena.nt.bioinf@gmail.com
github_repo: https://github.com/elenantbioinf/poc_cf
params:
    final_report_json: NULL
    target_region_evaluability_plot: NULL
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Montserrat:ital,wght@0,100..900;1,100..900&display=swap" rel="stylesheet">


```{css echo=FALSE}
.author, .title {
    display: none;
}

.main-container{
    max-width: 1000px;
    margin: 60px auto;
    padding: 50px;
    background-color: white;
    border: 1px solid #dcdcdc;
    border-radius: 10px;
    box-shadow: 0 4px 12px rgba(0,0,0,0.05);
}

.report-subtitle{
  text-align: center;
  margin-top: -25px;   /* lo acerca al título */
  margin-bottom: 30px; /* separa del resto */
  color: #333333;
  font-size: 16px;
  font-weight: 400;
}

.sample_information{
    font-size: 22px;
    line-height: 1.6;
}

h1 {
    text-align: center;
    background-color: #dedede;
    padding: 25px;
    border-bottom: 3px solid #1a1a1a;
    margin-bottom: 40px;
    border-radius: 6px;
}

h2 {
    padding: 25px;
    border-bottom: 3px solid #1a1a1a;
    margin-bottom: 40px;
    border-radius: 6px;
}

body {
    background-color: white;
    color: black;
    font-family: "Montserrat", sans-serif;
    font-size: 18px;
}
pre {
    background-color: #f4f4f4;
    padding: 15px;
    border-radius: 6px;
}
```

```{r echo=FALSE}
library(jsonlite)

data <- fromJSON(params$final_report_json)

sample_id <- data$sample_id
disease   <- data$disease
gene      <- data$gene
reference    <- data$reference_genome
target_region_evaluability_plot <- params$target_region_evaluability_plot
total_variants <- data$summary$total_unique_variants
variants <- data$variants
no_variants <- (total_variants == 0)
```

# CFTR Variant Report

<div class="report-subtitle">Proof-of-Concept report generated by Snakemake pipeline</div>

## Sample information

<div class="sample_information">

Sample ID: `r sample_id`

Disease: `r disease`

Gene: `r gene`

Reference genome: `r reference`

</div>

## Target region coverage evaluability

This figure summarizes which target BED intervals were technically evaluable according to the minimum coverage threshold used by the pipeline.

```{r target-region-evaluability, echo=FALSE, results="asis"}
if (!is.null(target_region_evaluability_plot) && file.exists(target_region_evaluability_plot)) {
  image_src <- knitr::image_uri(target_region_evaluability_plot)

  cat(
    sprintf(
      '<img src="%s" alt="Target region coverage evaluability map" style="width:100%%; display:block; margin:20px auto;">',
      image_src
    )
  )
} else {
  cat("Target region evaluability plot was not available.\n")
}
```

## Variants analysis

Total unique variants: `r total_variants`

```{r echo=FALSE, results="asis"}
if (no_variants) {
  cat("No variants detected in your analysis.\n")
} else {
  for (i in seq_len(nrow(variants))) {
    v <- variants[i, ]

    cat(sprintf("\n### Variant: %s\n\n", v$input_variant))
    cat(sprintf("- Best annotation: %s\n", v$best_annotation))
    cat(sprintf("- Gene: %s\n", v$gene))
    cat(sprintf("- Transcript: %s\n", v$transcript))
    cat(sprintf("- Exon: %s\n", v$exon))
    cat(sprintf("- Consequence: %s\n", v$consequence))
    cat(sprintf("- Impact: %s\n", v$impact))
    cat(sprintf("- HGVS_C: %s\n", v$hgvs_c))
    cat(sprintf("- Protein: %s\n", v$protein))
    cat(sprintf("- Transcript annotations (rows): %s\n\n", v$transcript_annotations_rows))
  }
}
```
