---
title: Analysis of CFTR variants
author: Elena Núñez Tomás
output: html_document
email_address: elena.nt.bioinf@gmail.com
github_repo: https://github.com/elenantbioinf/poc_cf
params:
    final_report_json: NULL
    target_region_evaluability_plot: NULL
    coverage_gaps_tsv: NULL
    gap_min_coverage: NULL
    target_bed: NULL
---

<link rel="preconnect" href="https://fonts.googleapis.com">
<link rel="preconnect" href="https://fonts.gstatic.com" crossorigin>
<link href="https://fonts.googleapis.com/css2?family=Montserrat:ital,wght@0,100..900;1,100..900&display=swap" rel="stylesheet">


```{css echo=FALSE}
.author, .title {
    display: none;
}

.main-container{
    max-width: 1000px;
    margin: 60px auto;
    padding: 50px;
    background-color: white;
    border: 1px solid #dcdcdc;
    border-radius: 10px;
    box-shadow: 0 4px 12px rgba(0,0,0,0.05);
}

.report-subtitle{
  text-align: center;
  margin-top: -25px;   /* lo acerca al título */
  margin-bottom: 30px; /* separa del resto */
  color: #333333;
  font-size: 16px;
  font-weight: 400;
}

.sample_information{
    font-size: 22px;
    line-height: 1.6;
}

h1 {
    text-align: center;
    background-color: #dedede;
    padding: 25px;
    border-bottom: 3px solid #1a1a1a;
    margin-bottom: 40px;
    border-radius: 6px;
}

h2 {
    padding: 25px;
    border-bottom: 3px solid #1a1a1a;
    margin-bottom: 40px;
    border-radius: 6px;
}

body {
    background-color: white;
    color: black;
    font-family: "Montserrat", sans-serif;
    font-size: 18px;
}
pre {
    background-color: #f4f4f4;
    padding: 15px;
    border-radius: 6px;
}

.coverage-summary {
    padding: 18px 22px;
    margin: 20px 0 35px 0;
    border-radius: 6px;
    background-color: #f7f7f7;
    font-size: 16px;
    line-height: 1.5;
}

.coverage-summary p {
    margin: 5px 0;
}

.coverage-ok {
    border-left: 5px solid #4d7f4d;
}

.coverage-warning {
    border-left: 5px solid #b2182b;
}
```

```{r echo=FALSE}
library(jsonlite)

data <- fromJSON(params$final_report_json)

sample_id <- data$sample_id
disease   <- data$disease
gene      <- data$gene
reference    <- data$reference_genome

target_bed <- params$target_bed
target_region_evaluability_plot <- params$target_region_evaluability_plot
coverage_gaps_tsv <- params$coverage_gaps_tsv
gap_min_coverage <- params$gap_min_coverage

total_variants <- data$summary$total_unique_variants
variants <- data$variants
no_variants <- (total_variants == 0)
```

# CFTR Variant Report

<div class="report-subtitle">Proof-of-Concept report generated by Snakemake pipeline</div>

## Sample information

<div class="sample_information">

Sample ID: `r sample_id`

Disease: `r disease`

Gene: `r gene`

Reference genome: `r reference`

</div>

## Target region coverage evaluability

This figure summarizes which target BED intervals were technically evaluable according to the minimum coverage threshold used by the pipeline.

```{r target-region-evaluability, echo=FALSE, results="asis"}
if (!is.null(target_region_evaluability_plot) && file.exists(target_region_evaluability_plot)) {
  image_src <- knitr::image_uri(target_region_evaluability_plot)

  cat(
    sprintf(
      '<img src="%s" alt="Target region coverage evaluability map" style="width:100%%; display:block; margin:20px auto;">',
      image_src
    )
  )
} else {
  cat("Target region evaluability plot was not available.\n")
}
```

### Coverage gap summary

```{r coverage-gap-summary, echo=FALSE, results="asis"}
format_bp <- function(value) {
  format(value, big.mark = ",", scientific = FALSE)
}

format_percent <- function(value) {
  sprintf("%.2f%%", value)
}

read_target_bed <- function(target_bed_file) {
  bed_df <- read.delim(
    target_bed_file,
    sep = "\t",
    header = FALSE,
    stringsAsFactors = FALSE,
    comment.char = "#"
  )

  if (ncol(bed_df) < 3) {
    stop("Target BED file must contain at least three columns: chrom, start, end.")
  }

  colnames(bed_df)[1:3] <- c("chrom", "start", "end")

  bed_df$start <- as.integer(bed_df$start)
  bed_df$end <- as.integer(bed_df$end)

  return(bed_df)
}

count_affected_target_intervals <- function(bed_df, gaps_df) {
  affected <- vapply(
    seq_len(nrow(bed_df)),
    function(index) {
      any(
        gaps_df$chrom == bed_df$chrom[index] &
          gaps_df$end > bed_df$start[index] &
          gaps_df$start < bed_df$end[index]
      )
    },
    logical(1)
  )

  sum(affected)
}

if (
  !is.null(coverage_gaps_tsv) && file.exists(coverage_gaps_tsv) &&
  !is.null(target_bed) && file.exists(target_bed)
) {

  gaps_df <- read.delim(
    coverage_gaps_tsv,
    sep = "\t",
    stringsAsFactors = FALSE
  )

  bed_df <- read_target_bed(target_bed)

  total_target_bp <- sum(bed_df$end - bed_df$start)
  total_target_intervals <- nrow(bed_df)

  if (nrow(gaps_df) == 0) {

    cat(
      paste0(
        "\n<div class='coverage-summary coverage-ok'>",
        "<p>No low-coverage gaps were detected below ", gap_min_coverage, "X.</p>",
        "<p>All target BED intervals are technically evaluable according to this coverage criterion.</p>",
        "<p>Total target length assessed: ", format_bp(total_target_bp), " bp.</p>",
        "</div>\n"
      )
    )

  } else {

    gaps_df$start <- as.integer(gaps_df$start)
    gaps_df$end <- as.integer(gaps_df$end)
    gaps_df$length <- as.integer(gaps_df$length)

    total_low_coverage_bp <- sum(gaps_df$length, na.rm = TRUE)
    low_coverage_bp_percent <- (total_low_coverage_bp / total_target_bp) * 100

    affected_target_intervals <- count_affected_target_intervals(
      bed_df = bed_df,
      gaps_df = gaps_df
    )

    affected_target_intervals_percent <- (
      affected_target_intervals / total_target_intervals
    ) * 100

    cat(
      paste0(
        "\n<div class='coverage-summary coverage-warning'>",
        "<p>Low-coverage gaps detected: ", nrow(gaps_df), "</p>",
        "<p>Minimum coverage threshold: ", gap_min_coverage, "X</p>",
        "<p>Total target length assessed: ", format_bp(total_target_bp), " bp</p>",
        "<p>Total low-coverage length: ", format_bp(total_low_coverage_bp), " bp</p>",
        "<p>Non-evaluable target bases: ", format_percent(low_coverage_bp_percent), "</p>",
        "<p>Target intervals affected by low coverage: ", affected_target_intervals,
        " / ", total_target_intervals,
        " (", format_percent(affected_target_intervals_percent), ")</p>",
        "<p>Regions overlapping these gaps should be considered technically non-evaluable for variant interpretation at this threshold.</p>",
        "</div>\n"
      )
    )
  }

} else {
  cat("Coverage gap summary was not available.\n")
}
```

## Variants analysis

Total unique variants: `r total_variants`

```{r echo=FALSE, results="asis"}
if (no_variants) {
  cat("No variants detected in your analysis.\n")
} else {
  for (i in seq_len(nrow(variants))) {
    v <- variants[i, ]

    cat(sprintf("\n### Variant: %s\n\n", v$input_variant))
    cat(sprintf("- Best annotation: %s\n", v$best_annotation))
    cat(sprintf("- Gene: %s\n", v$gene))
    cat(sprintf("- Transcript: %s\n", v$transcript))
    cat(sprintf("- Exon: %s\n", v$exon))
    cat(sprintf("- Consequence: %s\n", v$consequence))
    cat(sprintf("- Impact: %s\n", v$impact))
    cat(sprintf("- HGVS_C: %s\n", v$hgvs_c))
    cat(sprintf("- Protein: %s\n", v$protein))
    cat(sprintf("- Transcript annotations (rows): %s\n\n", v$transcript_annotations_rows))
  }
}
```
